---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# Water Point Functionality and Service Delivery Monitoring Dataset - 2023

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

This dataset contains water point assessment data collected during
routine field monitoring activities conducted in Malawi in 2023 under
the Trocaire water point functionality survey program. The survey was
designed to evaluate the operational status, service reliability, and
management conditions of rural water supply infrastructure, with a focus
on boreholes and community-managed water points.

Data collection was carried out using the mWater platform. The dataset
includes geospatial coordinates, functionality status, water
availability and reliability indicators, pump and civil works condition,
reported mechanical failures, seasonal variability, water quality
observations (including salinity and contamination issues), and basic
performance metrics such as collection time and access distance.

The dataset also captures governance and service delivery information,
including water point committee composition, tariff structures,
financial record keeping, maintenance practices, and service provider
arrangements.

The survey was conducted across multiple rural water points in Malawi,
with locations determined by program implementation and field
accessibility.

This dataset can be used for:

1.  Rural water infrastructure functionality analysis

2.  Borehole and handpump performance monitoring

3.  WASH program evaluation and reporting

4.  Geospatial mapping of service delivery gaps

5.  Water quality risk screening (e.g., salinity and contamination
    indicators)

6.  Governance and sustainability assessments of community-managed water
    systems

## Installation

You can install the development version of waterpointstatus from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/waterpointstatus")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/waterpointstatus/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to water point assessment data collected during routine field monitoring activities conducted in Malawi in 2023 under the Trocaire water point functionality survey program. The survey was designed to evaluate the operational status, service reliability, and management conditions of rural water supply infrastructure, with a focus on boreholes and community-managed water points.

```{r}
library(waterpointstatus)
```

### metadata

The dataset `waterpointstatus` contains `r nrow(waterpointstatus)`
observations and `r ncol(waterpointstatus)` variables

```{r}
waterpointstatus |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "waterpointstatus.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
library(waterpointstatus)

# Example 1: Visualization Map on water point status
library(leaflet)
library(dplyr)
library(stringr)

# -----------------------------
# 1. Data preparation
# -----------------------------
wp_map <- waterpointstatus %>%
  mutate(
    latitude = as.numeric(latitude),
    longitude = as.numeric(longitude),

    # Clean functional status (real-world messy labels handled)
    functional_status = case_when(
      str_detect(tolower(functional_status), "functional") &
        !str_detect(tolower(functional_status), "not|non") ~ "Functional",

      str_detect(tolower(functional_status), "not|non") ~ "Non-functional",

      TRUE ~ "Unknown / Partial"
    ),

    # Population
    user_population = as.numeric(user_population),

    # Scale for visualization
    pop_scaled = sqrt(user_population)
  ) %>%
  filter(!is.na(latitude), !is.na(longitude))

# -----------------------------
# 2. Color palette
# -----------------------------
status_palette <- colorFactor(
  palette = c(
    "Functional" = "green",
    "Non-functional" = "red",
    "Unknown / Partial" = "orange"
  ),
  domain = wp_map$functional_status
)

# -----------------------------
# 3. Leaflet map
# -----------------------------
leaflet(wp_map) %>%
  addProviderTiles(providers$OpenStreetMap) %>%

  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,

    color = ~status_palette(functional_status),
    fillColor = ~status_palette(functional_status),

    fillOpacity = 0.75,
    stroke = TRUE,
    weight = 1,

    radius = ~ifelse(is.na(pop_scaled), 4, pop_scaled / 6),

    popup = ~paste0(
      "<b>Water Point ID:</b> ", waterpoint_id, "<br>",
      "<b>Status:</b> ", functional_status, "<br>",
      "<b>User population:</b> ", ifelse(is.na(user_population), "NA", user_population), "<br>",
      "<b>Fill time (20L):</b> ", ifelse(is.na(fill_time_20l), "NA", fill_time_20l), "<br>",
      "<b>Repair days:</b> ", ifelse(is.na(repair_duration_days), "NA", repair_duration_days), "<br>",
      "<b>Latrine within 100m:</b> ", latrine_within_100m
    )
  ) %>%

  addLegend(
    position = "bottomright",
    pal = status_palette,
    values = ~functional_status,
    title = "Water Point Status"
  )
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("waterpointstatus")
```
